---
title: "ECHO_disparity-analyses"
author: "Travis Park, Natalie Cooper, Thomas Guillerme"
date: "16/02/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ECHO: Disparity analyses

This document takes the PC scores obtained fom the PCAs in the ECHO_GPA-ACE document and uses them to run a series of disparity analyses, including: disparity-through-time; between group disparity; and time-slicing. 

Remember to alter file paths to match your own directories.

## Load libraries

```{r}
library(here)
library(devtools)
library(ape)
library(Claddis)
library(dispRity)
library(dplyr)
library(broom)
library(tibble)
library(jtools)
library(huxtable)
```

## Load tree

```{r}
ECHO_tree <- read.tree(here("Data/BAMM_testtree_4.tre"))
# Adding node labels
ECHO_tree <- makeNodeLabel(ECHO_tree, method = "number", prefix = "n")
# Adding a root age (max tree age date) as write.nexus/tree won't save root times
ECHO_tree$root.time <- max(tree.age(ECHO_tree)[, 1])
```

## Load PC scores files

```{r}
## PC scores with ACE
ECHO_ace_pca_matrix <-  read.csv(here("Data/ECHO_ace_pca-scores_2.csv"))
# Make the taxon names the row names and then remove the taxon column from the data
row.names(ECHO_ace_pca_matrix) <- ECHO_ace_pca_matrix$X
ECHO_ace_pca_matrix <- select(ECHO_ace_pca_matrix, -X)
# format the csv file as a matrix
ECHO_ace_pca_matrix <- as.matrix(ECHO_ace_pca_matrix)
```

## Set jtools values
This sets the number of decimal points displayed when using the jtool function 'summ', which will be used below to combine our lm results into a table:

```{r}
set_summ_defaults(digits = 5)
```

## Disparity-through-time analyses

This analysis will measure changes in disparity through time by separating the data (can be specimens and/or the calculated ancestral states) into time bins (which can be defined manually or set from geological time divisions), where disparity is then calculated for each time bin. We then test for differences between the time bins.

# Load in geological time bin ages

These come from the International Commission on Stratigraphy 2015 ages:

```{r}
strat_ages <- get.bin.ages(ECHO_tree, type = "Epoch", ICS = 2015)
```

We will now rarefy the original data to remove sampling biases.

To run these analyses the standard 'dispRity.through.time' wrapper cannot be used as it doesn't let you include the node values. Instead you have to run the analysis step by step, as shown below.

Sum of variances (the sum of the variances of each axis of the matrix):
```{r}
# Delineate your time bins and indicate that nodes are also to be included
aceDTT_sumvar_rar <- chrono.subsets(ECHO_ace_pca_matrix, ECHO_tree, "discrete", time = strat_ages, inc.nodes = TRUE)
# Rarefy the matrix
aceDTT_sumvar_rar <- boot.matrix(aceDTT_sumvar_rar, rarefaction = "min")
# calculate the disparity values for your time bins
aceDTT_sumvar_rar <- dispRity(aceDTT_sumvar_rar, metric = c(sum, variances))
# Summarise DTT results
aceDTT_sum1 <- summary.dispRity(aceDTT_sumvar_rar)
aceDTT_sum1 <- add_column(aceDTT_sum1, 
             analysis = "DTT", 
             metric = "sumvar", 
             model = aceDTT_sumvar_rar$call$subsets[2],
             .before = 1
             )
# export  as csv file
write.csv(aceDTT_sum1, (here("Outputs/aceDTT_results.csv")))
# Basic plot of DTT results
plot.dispRity(aceDTT_sumvar_rar, type = "continuous", main = "aceDTT_sumvar_rar")
# Test for differences between the time bins using a linear model
aceDTT_sumvar_rar_lm <- test.dispRity(aceDTT_sumvar_rar, test = lm, comparisons = "all", correction = "bonferroni")
# Summarise lm results
aceDTT_lm_sum1 <- summ(aceDTT_sumvar_rar_lm)
# export lm summary as spreadsheet 
export_summs(aceDTT_lm_sum1, statistics = "all", to.file = "xlsx", file.name = (here("Outputs/aceDTT_lm_results.xlsx")))
```

## Time-slicing analysis

The ‘time-slicing’ approach considers subsets of taxa in the morphospace at specific equidistant points in time, as opposed to considering subsets of taxa  between two points in time. This results in even-sampling of the morphospace across time and allows us to use different underlying models of character evolution (punctuated or gradual). In practice, time-slicing considers the disparity of any element present in the phylogeny (branches, nodes and tips) at any point in time. When the phylogenetic elements are nodes or tips, the ordination scores for the nodes or tips are directly used for calculating disparity. When the phylogenetic elements are branches we choose the ordination score for the branch using one of two evolutionary models, which can be implemented in 6 different ways. These 6 models are all implemented using the sum of variances disparity metric for both 1 Ma and 5 Ma time slices, resulting in 12 separate analyses. We also test the metric below too, to ensure it is a suitable metric (i.e. a good proxy of tait space size). Please also note that all these analyses are rarefied.

### 1 Ma timeslice

# Sum of variances

Sum of variances metric, acctran model:
```{r}
# Creating the timeslices
ts1_sumvar_acctran_rar <- chrono.subsets(ECHO_ace_pca_matrix, ECHO_tree, method = "continuous", model = "acctran", time = 36, inc.nodes = TRUE, t0 = 35)
# Bootstrap the analysis
ts1_sumvar_acctran_rar <- boot.matrix(ts1_sumvar_acctran_rar, rarefaction = "min")
# Calculate disparity values
ts1_sumvar_acctran_rar <- dispRity(ts1_sumvar_acctran_rar, metric = c(sum, variances))
# Summarise TS results
ts1_sum1 <- summary.dispRity(ts1_sumvar_acctran_rar)
ts1_sum1 <- add_column(ts1_sum1, 
             analysis = "ts1", 
             metric = "sumvar", 
             model = ts1_sumvar_acctran_rar$call$subsets[2],
             .before = 1
             )
# Basic plot of TS results
plot.dispRity(ts1_sumvar_acctran_rar, type = "continuous", main = "ts1_sumvar_acctran_rar")
# Test for differences between the time slices using a linear model
ts1_sumvar_acctran_rar_lm <- test.dispRity(ts1_sumvar_acctran_rar, test = lm, comparisons = "all", correction = "bonferroni")
# Summarise lm results
ts1_lm_sum1 <- summ(ts_sumvar_acctran_rar_lm)
```

Sum of variances metric, deltran model:
```{r}
# Creating the timeslices
ts1_sumvar_deltran_rar <- chrono.subsets(ECHO_ace_pca_matrix, ECHO_tree, method = "continuous", model = "deltran", time = 36, inc.nodes = TRUE, t0 = 35)
# Bootstrap the analysis
ts1_sumvar_deltran_rar <- boot.matrix(ts1_sumvar_deltran_rar, rarefaction = "min")
# Calculate disparity values
ts1_sumvar_deltran_rar <- dispRity(ts1_sumvar_deltran_rar, metric = c(sum, variances))
# Summarise TS results
ts1_sum2 <- summary.dispRity(ts1_sumvar_deltran_rar)
ts1_sum2 <- add_column(ts1_sum2, 
             analysis = "ts1", 
             metric = "sumvar", 
             model = ts1_sumvar_deltran_rar$call$subsets[2],
             .before = 1
             )
# Basic plot of TS results
plot.dispRity(ts1_sumvar_deltran_rar, type = "continuous", main = "ts1_sumvar_deltran_rar")
# Test for differences between the time slices using a linear model
ts1_sumvar_deltran_rar_lm <- test.dispRity(ts1_sumvar_deltran_rar, test = lm, comparisons = "all", correction = "bonferroni")
# Summarise lm results
ts1_lm_sum2 <- summ(ts1_sumvar_deltran_rar_lm)
```

Sum of variances metric, random model:
```{r}
# Creating the timeslices
ts1_sumvar_rand_rar <- chrono.subsets(ECHO_ace_pca_matrix, ECHO_tree, method = "continuous", model = "random", time = 36, inc.nodes = TRUE, t0 = 35)
# Bootstrap the analysis
ts1_sumvar_rand_rar <- boot.matrix(ts1_sumvar_rand_rar, rarefaction = "min")
# Calculate disparity values
ts1_sumvar_rand_rar <- dispRity(ts1_sumvar_rand_rar, metric = c(sum, variances))
# Summarise TS results
ts1_sum3 <- summary.dispRity(ts1_sumvar_rand_rar)
ts1_sum3 <- add_column(ts1_sum3, 
             analysis = "ts1", 
             metric = "sumvar", 
             model = ts1_sumvar_rand_rar$call$subsets[2],
             .before = 1
             )
# Basic plot of TS results
plot.dispRity(ts1_sumvar_rand_rar, type = "continuous", main = "ts1_sumvar_rand_rar")
# Test for differences between the time slices using a linear model
ts1_sumvar_rand_rar_lm <- test.dispRity(ts1_sumvar_rand_rar, test = lm, comparisons = "all", correction = "bonferroni")
# Summarise lm results
ts1_lm_sum3 <- summ(ts1_sumvar_rand_rar_lm)
```

Sum of variances metric, proximity model:
```{r}
# Creating the timeslices
ts1_sumvar_prox_rar <- chrono.subsets(ECHO_ace_pca_matrix, ECHO_tree, method = "continuous", model = "proximity", time = 36, inc.nodes = TRUE, t0 = 35)
# Bootstrap the analysis
ts1_sumvar_prox_rar <- boot.matrix(ts1_sumvar_prox_rar, rarefaction = "min")
# Calculate disparity values
ts1_sumvar_prox_rar <- dispRity(ts1_sumvar_prox_rar, metric = c(sum, variances))
# Summarise TS results
ts1_sum4 <- summary.dispRity(ts1_sumvar_prox_rar)
ts1_sum4 <- add_column(ts1_sum4, 
             analysis = "ts1", 
             metric = "sumvar", 
             model = ts1_sumvar_prox_rar$call$subsets[2],
             .before = 1
             )
# Basic plot of TS results
plot.dispRity(ts1_sumvar_prox_rar, type = "continuous", main = "ts1_sumvar_prox_rar")
# Test for differences between the time slices using a linear model
ts1_sumvar_prox_rar_lm <- test.dispRity(ts1_sumvar_prox_rar, test = lm, comparisons = "all", correction = "bonferroni")
# Summarise lm results
ts1_lm_sum4 <- summ(ts1_sumvar_prox_rar_lm)
```

Sum of variances metric, equal split model:
```{r}
# Creating the timeslices
ts1_sumvar_eqsplit_rar <- chrono.subsets(ECHO_ace_pca_matrix, ECHO_tree, method = "continuous", model = "equal.split", time = 36, inc.nodes = TRUE, t0 = 35)
# Bootstrap the analysis
ts1_sumvar_eqsplit_rar <- boot.matrix(ts1_sumvar_eqsplit_rar, rarefaction = "min")
# Calculate disparity values
ts1_sumvar_eqsplit_rar <- dispRity(ts1_sumvar_eqsplit_rar, metric = c(sum, variances))
# Summarise TS results
ts1_sum5 <- summary.dispRity(ts1_sumvar_eqsplit_rar)
ts1_sum5 <- add_column(ts1_sum5, 
             analysis = "ts1", 
             metric = "sumvar", 
             model = ts1_sumvar_eqsplit_rar$call$subsets[2],
             .before = 1
             )
# Basic plot of TS results
plot.dispRity(ts1_sumvar_eqsplit_rar, type = "continuous", main = "ts1_sumvar_eqsplit_rar")
# Test for differences between the time slices using a linear model
ts1_sumvar_eqsplit_rar_lm <- test.dispRity(ts1_sumvar_eqsplit_rar, test = lm, comparisons = "all", correction = "bonferroni")
# Summarise lm results
ts1_lm_sum5 <- summ(ts1_sumvar_eqsplit_rar_lm)
```

Sum of variances metric, gradual split model:
```{r}
# Creating the timeslices
ts1_sumvar_grdsplit_rar <- chrono.subsets(ECHO_ace_pca_matrix, ECHO_tree, method = "continuous", model = "gradual.split", time = 36, inc.nodes = TRUE, t0 = 35)
# Bootstrap the analysis
ts1_sumvar_grdsplit_rar <- boot.matrix(ts1_sumvar_grdsplit_rar, rarefaction = "min")
# Calculate disparity values
ts1_sumvar_grdsplit_rar <- dispRity(ts1_sumvar_grdsplit_rar, metric = c(sum, variances))
# Summarise TS results
ts1_sum6 <- summary.dispRity(ts1_sumvar_grdsplit_rar)
ts1_sum6 <- add_column(ts1_sum6, 
             analysis = "ts1", 
             metric = "sumvar", 
             model = ts1_sumvar_grdsplit_rar$call$subsets[2],
             .before = 1
             )
# Basic plot of TS results
plot.dispRity(ts1_sumvar_grdsplit_rar, type = "continuous", main = "ts1_sumvar_grdsplit_rar")
# Test for differences between the time slices using a linear model
ts1_sumvar_grdsplit_rar_lm <- test.dispRity(ts1_sumvar_grdsplit_rar, test = lm, comparisons = "all", correction = "bonferroni")
# Summarise lm results
ts1_lm_sum6 <- summ(ts1_sumvar_grdsplit_rar_lm)
```

## Create dataframe combining all ts1 results
```{r}
# time slice analyses
# ts1 summaries list
ts1_summary_list <- list(ts1_sum1, ts1_sum2, ts1_sum3, ts1_sum4, ts1_sum5, ts1_sum6)
# combine ts1 summaries into singel dataframe
ts1_results <- bind_rows(ts1_summary_list)
# export dataframe as csv file
write.csv(ts1_results, (here("Outputs/ts1_results.csv")))

#lm analyses 
#ts1 lm summaries list
ts1_lm_summary_list <- list(ts1_lm_sum1, ts1_lm_sum2,ts1_lm_sum3, ts1_lm_sum4, ts1_lm_sum5, ts1_lm_sum6)
# model name list
ts1_lm_model_list <- list("ts1_sumvar_acctran", "ts1_sumvar_deltran", "ts1_sumvar_rand", "ts1_sumvar_prox", "ts1_sumvar_eqsplit", "ts1_sumvar_grdsplit")
# combine lm summaries into single table and export as spreadsheet 
ts1_lm_results <- export_summs(ts1_lm_summary_list, statistics = "all", model.names = ts1_lm_model_list, to.file = "xlsx", file.name = (here("Outputs/ts1_lm_results.xlsx")))
```

## Testing the diaparity metric

Finally, we will also test our chosen disparity metric, the sum of variances. In this study we are essentially using it as a proxy of trait space size (i.e. cochlear morphospace size). The code below will create a plot to show that this metric is indeed capturing changes in trait space size:

```{r}
sumvar_test <- test.metric(ECHO_ace_pca_matrix, metric = c(sum, variances), shifts = c("random", "size"), replicates = 50, save.steps = TRUE, verbose = TRUE)

plot(sumvar_test)
```